---
title: "One-to-One Table Matching Without Shared Keys; the Argeninean subset."
author: "Biodiversity Data Integration"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
geometry: margin=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

Aquí se documenta cómo se vincularon los comentarios y revisiones de Daniela Milano y Juan Pablo Barriga sobre las ocurrencias reportadas en territorio Argentino.  Se implementó la misma aproximación descrita para el total del set de datos (ver `Matching_full_dedup_dataset_with_comments_2025-12-06.Rmd`).   

Este documento describe el método empleado para restablecer el vínculo uno-a-uno entre la tabla de comentarios y revisiones (table1) y la de la base de datos (table2) de modo de poder ejecutar dichos cambios, asegurando correspondencias precisas y completamente trazables.  

**Principios clave del flujo de trabajo**:  

  - Cada fila de `table1` debe coincidir de manera exacta con una única fila de `table2.`  
  - Las comparaciones se realizan exhaustivamente contra todas las filas de la otra tabla.  
  - Las coincidencias se establecen mediante combinaciones incrementales de campos clave previamente armonizados.  
  - El orden estratégico de estos campos favorece coincidencias correctas, conservadoras y no ambiguas.   


# Load Required Libraries

```{r libraries}

setwd("C:/Users/crist/Documents/GitHub/PEW-ProAp_Atlas Ictiogeografico_2025/Atlas_Ictio_project/")
rm(list=ls())


library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
library(tidyr)
```


<!-- En esta tabla se excluyeron algunos bloques de datos: -->
<!--   - (DatasetName == GBIF) & (institutionCode == iNaturalist). Ahora los datos de iNat tienen precedencia, y además se excluyenos los datos de iNat vía GBIF.   -->
<!--   - (datasetName == "gbif") & (str_sub(occurrenceID, 1L, 3L) == "mma"). Usamos la versión mejorada directamente del MMA, no la versión  publicada en GBIF. -->
<!--   - (datasetName == "subpesca-1") una lista de papers indexados en otras funtes fue excluido. Ver planilla en la respectiva carpeta,  y código de exclusión arriba.   -->


# Data importation    

Current database  
```{r data-entry}
# Load data to start from here 

X.filtered.dedup <- readRDS("./input/dataset_before_Remocion_manual_2025-12-03.rds")


```


Database version submitted to Daniela Milano and Juan Pablo Barriga   

```{r}
DB.submitted <- read.csv("./input/AtlasIctio2025_Datos_DwC_deInteres_norepetidos.csv", sep = ",", colClasses = "character", na.strings = "", fileEncoding = "UTF-8")


# Merge (coalesce) two columns that were separated only due to a typo.

DB.submitted <- DB.submitted %>%
  mutate(
    across(c(bibliographicCitation, bibiographicCitation),
           ~ str_trim(.) %>% na_if("") %>% na_if("NA")),
    bibliographicCitation = coalesce(bibliographicCitation, bibiographicCitation)
  ) %>%  select(-bibiographicCitation)   
```



Database received back with comments   
```{r}

# CM.Ar <- read.csv("./input_parameters/AtlasIctio_COMMENTS_DMilano_JPBarriga_Argentina.csv", sep = ",", colClasses = "character",stringsAsFactors = F, na.strings = "", fileEncoding = "UTF-8")


CM.Ar <- readxl::read_xlsx("./input/AtlasIctio2025_Datos_DwC_deInteres_norepetidos DM.xlsx", na = c("NA", ""), col_types = "text")

CM.Ar <- CM.Ar %>%
  mutate(eventDate = if_else(
    grepl("^[0-9]+$", eventDate),  # only digits → likely Excel serial
    as.character(as.Date(as.numeric(eventDate), origin = "1900-01-01")), #origin = "1899-12-30"
    eventDate
  ))

## Fix character encoding-related issues. 
CM.Ar <- CM.Ar %>%  
  mutate(across(
    .cols = where(is.character) & 
            !any_of("correccion") & 
            !starts_with("."),
    .fns = ~ iconv(., from = "UTF-8", to = "Windows-1252", sub = "byte")
  ))

# CM.Ar %>% select(locality, correccion, starts_with(".")) %>% filter(!is.na(correccion)) 


# Merge (coalesce) two columns that were separated only due to a typo.

CM.Ar <- CM.Ar %>%
  mutate(
    across(c(bibliographicCitation, bibiographicCitation),
           ~ str_trim(.) %>% na_if("") %>% na_if("NA")),
    bibliographicCitation = coalesce(bibliographicCitation, bibiographicCitation)
  ) %>%  select(-bibiographicCitation)


# Remove "NUEVOS" and sort by id_CC
CM.Ar <- CM.Ar %>% filter(id_CC != "NUEVO") %>% 
   arrange(as.numeric(id_CC))

# Rename the correccion column to make it compatible with code below 
CM.Ar <- CM.Ar %>% mutate(Comment = correccion) %>%
  # relocate(Comment, .before = .duplicate) %>% 
  select(-correccion)

# Keep only rows with comments in another object
edits.Ar <- CM.Ar %>% 
  # filter(rowSums(!is.na(across(c(Comment, starts_with(".")))) &
  #           across(c(Comment, starts_with(".")), ~ .x != "")) > 0  )
  filter(!is.na(Comment))

```


<!-- **HERE** Trying to match CM.Ar with CM (table1_raw) failed. table1_raw and CM were loaded from the other script  -->  

```{r, eval = T }

baseline <- DB.submitted #%>% select(-starts_with("."), -Duplicate)

edits.Ar.matched <- baseline %>% right_join(edits.Ar, by = "id_CC", keep = F) %>% 
  select(id_CC, starts_with("canonicalName"), Comment) 

# LOTS OF ROWS IN Ar WERE NOT FOUND IN CM 
# rm(edits.Ar.fixed, edits.Ar, baseline, edits.Ar.matched)

cat("The proportion of missmatching canonicalName is:",
    100 * sum(edits.Ar.matched$canonicalName.x != edits.Ar.matched$canonicalName.y, na.rm = T) /  
    nrow(edits.Ar.matched), "%" )

```
# Join the submitted and the received table by id_CC, and compare corresponding fields  between both tables

Note: Self-comparisson produces perfect or almost perfect match in datasetName, coordinates, canonicalName, locality and bibliographicCitation. 

```{r}

assess_field_correspondence <- function(baseline, edits, join_key = "id_CC", 
                                       candidate_fields, plot = TRUE) {
  
  # Join tables
  joined <- baseline %>% 
    right_join(edits, by = join_key, suffix = c(".baseline", ".edits"), keep = FALSE)
  
  # Initialize results data frame
  results <- data.frame(
    field = character(),
    n_total = integer(),
    n_matches = integer(),
    n_mismatches = integer(),
    n_both_na = integer(),
    n_only_baseline_na = integer(),
    n_only_edits_na = integer(),
    pct_matches = numeric(),
    pct_mismatches = numeric(),
    pct_any_na = numeric(),
    stringsAsFactors = FALSE
  )
  
  # Assess each field
  for (field in candidate_fields) {
    
    col_baseline <- paste0(field, ".baseline")
    col_edits <- paste0(field, ".edits")
    
    # Check if columns exist
    if (!col_baseline %in% names(joined) || !col_edits %in% names(joined)) {
      warning(paste("Field", field, "not found in one or both tables. Skipping."))
      next
    }
    
    baseline_vals <- joined[[col_baseline]]
    edits_vals <- joined[[col_edits]]
    
    # Calculate statistics
    n_total <- nrow(joined)
    
    both_na <- is.na(baseline_vals) & is.na(edits_vals)
    only_baseline_na <- is.na(baseline_vals) & !is.na(edits_vals)
    only_edits_na <- !is.na(baseline_vals) & is.na(edits_vals)
    both_present <- !is.na(baseline_vals) & !is.na(edits_vals)
    
    matches <- both_present & (baseline_vals == edits_vals)
    mismatches <- both_present & (baseline_vals != edits_vals)
    
    n_matches <- sum(matches, na.rm = TRUE)
    n_mismatches <- sum(mismatches, na.rm = TRUE)
    n_both_na <- sum(both_na, na.rm = TRUE)
    n_only_baseline_na <- sum(only_baseline_na, na.rm = TRUE)
    n_only_edits_na <- sum(only_edits_na, na.rm = TRUE)
    
    # Add to results
    results <- rbind(results, data.frame(
      field = field,
      n_total = n_total,
      n_matches = n_matches,
      n_mismatches = n_mismatches,
      n_both_na = n_both_na,
      n_only_baseline_na = n_only_baseline_na,
      n_only_edits_na = n_only_edits_na,
      pct_matches = round(100 * n_matches / n_total, 1),
      pct_mismatches = round(100 * n_mismatches / n_total, 1),
      pct_any_na = round(100 * (n_both_na + n_only_baseline_na + n_only_edits_na) / n_total, 1),
      stringsAsFactors = FALSE
    ))
  }
  
  # Sort by mismatch rate
  results <- results %>% arrange(desc(pct_mismatches), pct_matches)
  
  # Create plot if requested
  if (plot && nrow(results) > 0) {
    
    # Prepare data for plotting
    plot_data <- results %>%
      select(field, n_matches, n_mismatches, n_both_na, 
             n_only_baseline_na, n_only_edits_na) %>%
      pivot_longer(cols = -field, names_to = "category", values_to = "count") %>%
      mutate(
        category = factor(category, 
                         levels = c("n_matches", "n_mismatches", "n_both_na",
                                   "n_only_baseline_na", "n_only_edits_na"),
                         labels = c("Matches", "Mismatches", "Both NA",
                                   "Baseline NA", "Edits NA")),
        field = factor(field, levels = rev(results$field))
      )
    
    # Create stacked barplot
    p <- ggplot(plot_data, aes(x = field, y = count, fill = category)) +
      geom_bar(stat = "identity", position = "stack") +
      coord_flip() +
      scale_fill_manual(
        values = c("Matches" = "#2ecc71", 
                   "Mismatches" = "#e74c3c",
                   "Both NA" = "#95a5a6",
                   "Baseline NA" = "#f39c12",
                   "Edits NA" = "#e67e22")
      ) +
      labs(
        title = "Field Correspondence Assessment",
        subtitle = paste("Comparing", nrow(joined), "records"),
        x = "Field",
        y = "Number of Records",
        fill = "Status"
      ) +
      theme_minimal() +
      theme(
        legend.position = "bottom",
        axis.text.y = element_text(size = 10)
      )
    
    print(p)
  }
  
  # Return results
  list(
    summary = results,
    joined_data = joined
  )
}

# Usage example:
candidate_fields <- c("canonicalName", "decimalLatitude", "decimalLongitude", 
                     "year", "eventDate", "locality", 
                     # "samplingProtocol",  "individualCount", "catalogNumber", 
                     "datasetName", 
                     "bibliographicCitation", "institutionCode", "occurrenceID")

assessment <- assess_field_correspondence(
  baseline = DB.submitted,# DB.submitted,
  edits = edits.Ar[!(edits.Ar$institutionCode %in% "FO"),], # Exclude Fundacion Oga
  join_key = "id_CC",
  candidate_fields = candidate_fields,
  plot = TRUE
) 

# View summary table
kable(assessment$summary, caption = "Field Correspondence Summary")

# Access specific results
cat("\nFields with highest mismatch rates:\n")
assessment$summary %>% 
  arrange(desc(pct_mismatches)) %>% 
  select(field, pct_mismatches, n_mismatches) %>%
  head(5)


```



```{r, eval=FALSE}
CM <- CM %>% mutate(decimalLatitude = as.numeric(decimalLatitude), 
       decimalLongitude= as.numeric(decimalLongitude), year = as.numeric(year))
```


# Data Preparation

## Filter Source Data

Extract rows from the comments table (CM) that require action:

```{r prepare-table1}
# Table1: rows with changes or exclusion requests
# Duplicate: "0" = exclude, "1" = keep, NA = no action needed
# Columns with "." prefix contain proposed changes


table1_raw <- edits.Ar %>% 
                select(
                  id_CC,
                  # Duplicate,
                  decimalLatitude,
                  decimalLongitude,
                  canonicalName, 
                  year, 
                  eventDate, 
                  locality, 
                  samplingProtocol,
                  individualCount,
                  catalogNumber, 
                  datasetName,
                  bibliographicCitation,
                  institutionCode,
                  occurrenceID, 
                  Comment,
                  starts_with(".")
                )

cat("Table1 rows requiring action:", nrow(table1_raw), "\n")

```
Note: There are several datasets that are not carried forward because no changes need to be made. These are:  


```{r prepare-table2}
# Table2: consolidated master table after de-duplication
table2_raw <- DB.submitted %>%
  select(any_of(c("id_CC", colnames(table1_raw))))

cat("Table2 total rows:", nrow(table2_raw), "\n")

# Verify id_CC is unique
if (any(duplicated(table2_raw$id_CC))) {
  stop("ERROR: id_CC in table2 is not unique!")
}
```

## Standardization Function

Reduce meaningless variation in key fields:

```{r standardization-function}
standardize_for_matching <- function(df) {
  df %>%
    mutate(
      # Coordinates: ensure numeric (no rounding to preserve precision)
      decimalLatitude  = as.numeric(decimalLatitude),
      decimalLongitude = as.numeric(decimalLongitude),
      
      # Event date: extract YYYY-MM-DD format
      eventDate = if_else(
        is.na(eventDate),
        NA_character_,
        str_sub(as.character(eventDate), 1L, 10L)
      ),
      
      # Year: from year column or extract from eventDate
      year = {
        y1 <- suppressWarnings(as.integer(year))
        y2 <- suppressWarnings(as.integer(str_extract(eventDate, "\\d{4}")))
        coalesce(y1, y2)
      },
      
      # Bibliographic citation: first 20 chars (avoid minor variations)
      bibliographicCitation = str_sub(
        str_trim(as.character(bibliographicCitation)),
        1, 20
      ),
      
      # Text fields: trim, lowercase, standardize NAs
      across(
        c(canonicalName, locality, datasetName, catalogNumber,
          occurrenceID, institutionCode),
        ~ {
          x <- str_to_lower(str_trim(as.character(.x)))
          x[x %in% c("", "na", "n a", "n/a")] <- NA_character_
          x
        }
      ),
      
      # Dataset name: standardize known variations
      datasetName = case_when(
        datasetName %in% c("subpesca-1", "subpesca-2",
                           "fish database correa-boisjoly",
                           "fishgis_metadatabase_ccorrea") ~ datasetName,
        TRUE ~ str_sub(datasetName, 1L, 4L)
      )
    )
}
```

## Apply Standardization

```{r apply-standardization}
table1 <- standardize_for_matching(table1_raw)
table2 <- standardize_for_matching(table2_raw)

cat("Standardization complete.\n")
cat("Table1:", nrow(table1), "rows\n")
cat("Table2:", nrow(table2), "rows\n")
```

# Matching Strategy

## Define Matching Keys

Keys are ordered by reliability:

```{r define-keys}
# Core keys: most reliable identifiers
core_keys <- c(
  "datasetName",
  "canonicalName",
  "decimalLatitude",
  "decimalLongitude",
  "locality",
  "bibliographicCitation")

# Extra keys: used to resolve ambiguities
extra_keys <- c(
  "catalogNumber",
  "year",
  "locality",
  "samplingProtocol",
  "individualCount",
  "eventDate",
  "institutionCode",
  "occurrenceID")

all_keys <- c(core_keys, extra_keys)

cat("Matching will use up to", length(all_keys), "keys in hierarchical order\n")
cat("Core keys:", length(core_keys), "\n")
cat("Extra keys:", length(extra_keys), "\n")
```

## Strictly Conservative Matching Function

This function adds keys incrementally until a unique match is found. If no unique match is found, no assignment is  done, and the match_id field is left blank:

```{r matching-function}
match_tables_strict <- function(table1, table2, core_keys, extra_keys) {
  
  # Validate inputs
  if (!"id_CC" %in% names(table2)) {
    stop("table2 must contain unique identifier 'id_CC'")
  }
  
  if (any(duplicated(table2$id_CC))) {
    stop("id_CC in table2 must be unique")
  }
  
  # Initialize result
  result <- table1 %>%
    mutate(
      row_id = row_number(),
      match_id = NA_integer_,
      match_keys_used = NA_character_,
      n_keys_used = NA_integer_,
      n_candidates_final = NA_integer_,
      match_status = NA_character_
    )
  
  all_keys <- c(core_keys, extra_keys)
  n_core <- length(core_keys)
  
  # Process each row of table1
  for (i in seq_len(nrow(table1))) {
    
    row_i <- table1[i, , drop = FALSE]
    
    # Start with ALL rows from table2 (no tracking of matched rows)
    candidates <- table2
    
    keys_used <- character(0)
    keys_available <- character(0)
    
    # Identify which keys have non-NA values in this table1 row
    for (key in all_keys) {
      if (key %in% names(row_i) && !is.na(row_i[[key]])) {
        keys_available <- c(keys_available, key)
      }
    }
    
    # Incrementally add keys until unique match or keys exhausted
    for (key in keys_available) {
      
      val <- row_i[[key]]
      
      # Filter candidates: exact match on this key, non-NA in table2
      candidates <- candidates %>%
        filter(!is.na(.data[[key]]), .data[[key]] == val)
      
      keys_used <- c(keys_used, key)
      
      # Stop if no candidates remain
      if (nrow(candidates) == 0) {
        result$match_status[i] <- "no_match"
        result$n_candidates_final[i] <- 0
        break
      }
      
      # Stop if unique match achieved
      if (nrow(candidates) == 1) {
        result$match_id[i] <- candidates$id_CC[1]
        result$match_keys_used[i] <- paste(keys_used, collapse = ";")
        result$n_keys_used[i] <- length(keys_used)
        result$n_candidates_final[i] <- 1
        result$match_status[i] <- "matched"
        break
      }
    }
    
    # If we exhausted all keys without achieving uniqueness
    if (is.na(result$match_status[i]) && nrow(candidates) > 1) {
      result$match_status[i] <- "ambiguous"
      result$n_candidates_final[i] <- nrow(candidates)
      result$match_keys_used[i] <- paste(keys_used, collapse = ";")
      result$n_keys_used[i] <- length(keys_used)
    }
    
    # If no keys were available for this row
    if (length(keys_available) == 0) {
      result$match_status[i] <- "no_keys_available"
      result$n_candidates_final[i] <- nrow(table2)
    }
  }
  
  # Classify match quality based on which keys were used
  result <- result %>%
    mutate(
      n_core_keys_used = if_else(
        !is.na(match_keys_used),
        sapply(strsplit(match_keys_used, ";"), function(keys) {
          sum(keys %in% core_keys)
        }),
        NA_integer_
      ),
      match_quality = case_when(
        match_status != "matched" ~ match_status,
        n_core_keys_used == n_core ~ "excellent_all_core",
        n_core_keys_used >= 4 ~ "good_4plus_core",
        n_core_keys_used == 3 ~ "acceptable_3_core",
        n_core_keys_used == 2 ~ "weak_2_core",
        n_core_keys_used <= 1 ~ "very_weak_1_core",
        TRUE ~ "matched"
      )
    )
  
  result %>% select(-row_id)
}
```

## Execute Matching

```{r execute-matching}
cat("Starting strictly conservative matching...\n")
cat("Each table1 row will be matched independently against ALL table2 rows.\n\n")

table1_matched <- match_tables_strict(
  table1 = table1,
  table2 = table2,
  core_keys = core_keys,
  extra_keys = extra_keys
)

cat("Matching complete.\n")
```

# Results and Diagnostics

## Overall Matching Success

```{r matching-summary}
summary_stats <- table1_matched %>%
  summarise(
    total_rows = n(),
    matched = sum(match_status == "matched", na.rm = TRUE),
    ambiguous = sum(match_status == "ambiguous", na.rm = TRUE),
    no_match = sum(match_status == "no_match", na.rm = TRUE),
    no_keys = sum(match_status == "no_keys_available", na.rm = TRUE),
    match_rate = round(100 * matched / total_rows, 1)
  )

kable(summary_stats, caption = "Overall Matching Results")
```

## Match Status Distribution

```{r match-status}
status_summary <- table1_matched %>%
  count(match_status) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1)
  ) %>%
  arrange(desc(n))

kable(status_summary, caption = "Match Status Breakdown")
```

```{r status-plot, fig.cap="Distribution of Match Status"}
table1_matched %>%
  count(match_status) %>%
  mutate(match_status = forcats::fct_reorder(match_status, n)) %>%
  ggplot(aes(x = match_status, y = n, fill = match_status)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Match Status Distribution",
    x = "Status",
    y = "Number of Rows"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Match Quality for Successfully Matched Rows

```{r match-quality}
quality_summary <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_quality) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1)
  ) %>%
  arrange(desc(n))

kable(quality_summary, caption = "Match Quality Breakdown (Matched Rows Only)")
```

```{r quality-plot, fig.cap="Distribution of Match Quality"}
table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_quality) %>%
  mutate(match_quality = forcats::fct_reorder(match_quality, n)) %>%
  ggplot(aes(x = match_quality, y = n, fill = match_quality)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Match Quality Distribution",
    subtitle = "For successfully matched rows only",
    x = "Quality Category",
    y = "Number of Matches"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Keys Required for Matching

```{r keys-used}
keys_summary <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(n_keys_used, n_core_keys_used) %>%
  arrange(n_keys_used, desc(n_core_keys_used))

kable(keys_summary, caption = "Number of Keys Required to Achieve Uniqueness")
```

```{r keys-plot, fig.cap="Distribution of Keys Used"}
table1_matched %>%
  filter(match_status == "matched") %>%
  ggplot(aes(x = n_keys_used, fill = match_quality)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Number of Keys Required for Unique Match",
    x = "Number of Keys Used",
    y = "Frequency",
    fill = "Match Quality"
  ) +
  theme_minimal()
```

## Specific Key Combinations

Top 20 most common key combinations used:

```{r key-combinations}
key_combos <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_keys_used, sort = TRUE) %>%
  head(20)

kable(key_combos, caption = "Most Common Key Combinations")
```

## Core vs Extra Keys Usage

```{r core-extra-keys}
core_extra <- table1_matched %>%
  filter(match_status == "matched") %>%
  mutate(
    n_extra_keys_used = n_keys_used - n_core_keys_used
  ) %>%
  count(n_core_keys_used, n_extra_keys_used) %>%
  arrange(desc(n))

kable(core_extra, caption = "Core vs Extra Keys Usage")
```

```{r core-extra-plot, fig.cap="Core vs Extra Keys Scatter"}
table1_matched %>%
  filter(match_status == "matched") %>%
  mutate(n_extra_keys_used = n_keys_used - n_core_keys_used) %>%
  count(n_core_keys_used, n_extra_keys_used) %>%
  ggplot(aes(x = n_core_keys_used, y = n_extra_keys_used, size = n)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  scale_size_continuous(range = c(3, 15)) +
  labs(
    title = "Core vs Extra Keys Required",
    x = "Number of Core Keys Used",
    y = "Number of Extra Keys Used",
    size = "Frequency"
  ) +
  theme_minimal()
```

## Uniqueness Validation

Critical check: verify each match_id appears only once:

```{r uniqueness-check}
duplicate_matches <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_id, name = "frequency") %>%
  filter(frequency > 1)

if (nrow(duplicate_matches) > 0) {
  cat(" ERROR:", nrow(duplicate_matches), "table2 rows matched multiple times!\n")
  kable(head(duplicate_matches, 20), caption = "Duplicate Matches (CRITICAL ERROR)")
} else {
  cat(" SUCCESS: All matches are unique (one-to-one)\n")
  cat("Each match_id appears exactly once.\n")
}
```

```{r uniqueness-plot, fig.cap="Match ID Frequency Distribution"}
table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_id, name = "freq") %>%
  ggplot(aes(x = freq)) +
  geom_histogram(binwidth = 1, color = "black", fill = "steelblue") +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = 1.5, y = Inf, label = "All should be at 1", 
           vjust = 2, color = "red", fontface = "bold") +
  labs(
    title = "Distribution of match_id Frequency",
    subtitle = "Perfect one-to-one matching requires all frequencies = 1",
    x = "Times each match_id appears",
    y = "Count"
  ) +
  theme_minimal()
```

## Ambiguous Matches Analysis

Rows that had multiple candidates even after using all available keys:

```{r ambiguous-analysis}
if (sum(table1_matched$match_status == "ambiguous", na.rm = TRUE) > 0) {
  
  ambiguous_summary <- table1_matched %>%
    filter(match_status == "ambiguous") %>%
    group_by(n_candidates_final) %>%
    summarise(
      count = n(),
      avg_keys_used = round(mean(n_keys_used, na.rm = TRUE), 1)
    ) %>%
    arrange(desc(count))
  
  kable(ambiguous_summary, caption = "Ambiguous Matches: Number of Final Candidates")
  
  # Show examples of ambiguous matches
  ambiguous_examples <- table1_matched %>%
    filter(match_status == "ambiguous") %>%
    select(
      id_CC, datasetName, canonicalName, year, 
      n_candidates_final, n_keys_used, match_keys_used
    ) %>%
    head(10)
  
  kable(ambiguous_examples, caption = "Examples of Ambiguous Matches")
  
} else {
  cat("No ambiguous matches found.\n")
}
```

## No Match Analysis

Rows where no table2 row matched the criteria:

```{r no-match-analysis}
if (sum(table1_matched$match_status == "no_match", na.rm = TRUE) > 0) {
  
  no_match_summary <- table1_matched %>%
    filter(match_status == "no_match") %>%
    count(datasetName) %>%
    arrange(desc(n))
  
  kable(no_match_summary, caption = "No Match Cases by Dataset")
  
} else {
  cat("All rows either matched or were ambiguous.\n")
}
```

## Unmatched Rows Requiring Review

Exclude expected unmatched categories:


```{r unmatched-review}
unmatched_review <- table1_matched %>%
  filter(
     match_status != "matched",
        ) %>%
  select(
    id_CC,
    match_status,
    match_quality,
    n_candidates_final,
    n_keys_used,
    match_keys_used,
    datasetName,
    canonicalName,
    year,
    eventDate,
    locality,
    decimalLatitude,
    decimalLongitude,
    bibliographicCitation,
    catalogNumber,
    occurrenceID,
    institutionCode
  ) %>%
  arrange(match_status, datasetName, canonicalName, year)

cat("Unmatched rows requiring manual review:", nrow(unmatched_review), "\n\n")

unmatched_by_status <- unmatched_review %>%
  count(match_status, datasetName) %>%
  arrange(match_status, desc(n))

kable(unmatched_by_status, caption = "Unmatched Rows by Status and Dataset")
```

## Matched, rowwise field-by-Field Comparison  

```{r rowwise-field-comparison, eval=TRUE} 

if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {

# matched subsets
tb1 <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    match_id,
    any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
  ) %>%
  arrange(match_id)

tb2 <- table2 %>%
  filter(id_CC %in% tb1$match_id) %>%
  arrange(id_CC)

tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]

comparison_cols <- setdiff(intersect(names(tb1), names(tb2)), "match_id")

diff.strict <- tb1[, comparison_cols] == tb2[, comparison_cols]

tmp <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    id_CC_old = id_CC,
    match_id,
    match_keys_used,
    dataset = datasetName
  ) %>%
  arrange(match_id)

# logical match matrix -> data frame
diff.strict <- as.data.frame(diff.strict)

# add metadata for each matched row
diff.strict <- bind_cols(
  tmp[!is.na(tmp$match_id), ],  # keep only matched rows in tmp
  diff.strict
)

# compute number of matching fields row-wise
diff.strict <- diff.strict %>%
  mutate(fieldgood = rowSums(across(!c(id_CC_old, match_id, match_keys_used, dataset)), na.rm = TRUE))

# plot
ggplot(diff.strict, aes(x = dataset, y = fieldgood)) +
  geom_jitter(aes(color = fieldgood),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
                     limits= c(1, length(comparison_cols)),
                     expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Row-wise Agreement: Number of Fields Matched",
    x = NULL,
    y = "Count of matching fields"
  ) +
  theme_minimal()

} else {
  cat("No matched rows to compare.\n")
}

```

```{r rowwise-field-comparison2, eval=TRUE} 


 if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {

# matched subsets
tb1 <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    match_id,
    any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
  ) %>%
  arrange(match_id)

tb2 <- table2 %>%
  filter(id_CC %in% tb1$match_id) %>%
  arrange(id_CC)

tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]

comparison_cols <- setdiff(intersect(names(tb1), names(tb2)), "match_id")

# COUNT DIFFERENCES THIS TIME
diff.strict2 <- tb1[, comparison_cols] != tb2[, comparison_cols]

tmp <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    id_CC_old = id_CC,
    match_id,
    match_keys_used,
    dataset = datasetName
  ) %>%
  arrange(match_id)

# logical match matrix -> data frame
diff.strict2 <- as.data.frame(diff.strict2)

# add metadata for each matched row
diff.strict2 <- bind_cols(
  tmp[!is.na(tmp$match_id), ],  # keep only matched rows in tmp
  diff.strict2
)

# compute number of matching fields row-wise
diff.strict.falses <- diff.strict2 %>%
  mutate(fieldbad = rowSums(across(!c(id_CC_old, match_id, match_keys_used, dataset)), na.rm = TRUE))

# plot
ggplot(diff.strict.falses, aes(x = dataset, y = fieldbad)) +
  geom_jitter(aes(color = fieldbad),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
                     limits= c(1, length(comparison_cols)),
                     expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Row-wise Agreement: Number of Fields Matched",
    x = NULL,
    y = "Count of non-matching fields"
  ) +
  theme_minimal()

} else {
  cat("No matched rows to compare.\n")
}



```


```{r}

good.and.bad.matches <-  diff.strict %>% full_join(diff.strict.falses, by = "id_CC_old", keep = T) %>% 
  select(dataset.x, id_CC_old.x, match_id.x, fieldgood, fieldbad) %>% 
  mutate(logodd.ratio = log10((fieldgood + 1) / (fieldbad + 1)), 
         precent.good = 100 * (fieldgood /length(comparison_cols)),
         percent.bad = 100 *  (fieldbad /length(comparison_cols))) 

good.and.bad.matches %>% filter(fieldbad >= 3) %>% 
  kable(caption = "Matched rows with 3 or more missmatches.")


# write.csv(
#     good.and.bad.matches,
#     "./outputs_global_analyses/Table1_good-bad_matches_Argentina.csv",
#     fileEncoding = "UTF-8",
#     row.names = FALSE
#   )

```

```{r}
good.and.bad.matches %>% 
ggplot(aes(x = dataset.x, y = logodd.ratio)) +
  geom_jitter(aes(color = logodd.ratio),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  # scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
  #                    limits= c(1, length(comparison_cols)),
  #                    expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Row-wise Agreement: Log-Ratio of matched-to-unmatched fields",
    x = NULL,
    y = "Log of the odds"
  ) +
  theme_minimal()


```



## Matched, overall field-by-Field Comparison  

For successfully matched rows, compare original values:

```{r overall-field-comparison, eval=TRUE}
 if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  
  # Prepare matched subsets
  tb1 <- table1_matched %>%
    filter(match_status == "matched") %>%
    select(
      match_id,
      any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
    ) %>%
    arrange(match_id)
  
  tb2 <- table2 %>%
    filter(id_CC %in% tb1$match_id) %>%
    arrange(id_CC)
  
  # Ensure same order
  tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]
  
  # Compare field by field
  comparison_cols <- intersect(names(tb1), names(tb2))
  comparison_cols <- setdiff(comparison_cols, "match_id")
  
  agreement <- sapply(comparison_cols, function(col) {
    sum(tb1[[col]] == tb2[[col]], na.rm = TRUE)
  })
  
  total <- nrow(tb1)
  
  agreement_summary <- data.frame(
    field = comparison_cols,
    agreements = agreement,
    total = total,
    agreement_rate = round(100 * agreement / total, 1)
  ) %>%
    arrange(desc(agreement_rate))
  
  kable(agreement_summary, caption = "Field Agreement Rates for Matched Rows")
  
} else {
  cat("No matched rows to compare.\n")
}
```



```{r agreement-plot, eval=F, fig.cap="Field Agreement Rates"}
if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  agreement_summary %>%
    mutate(field = forcats::fct_reorder(field, agreement_rate)) %>%
    ggplot(aes(x = field, y = agreement_rate, fill = agreement_rate)) +
    geom_col() +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    coord_flip() +
    scale_fill_gradient(low = "orange", high = "darkgreen") +
    labs(
      title = "Field Agreement Rates",
      subtitle = "Comparison between matched table1 and table2 rows",
      x = "Field",
      y = "Agreement Rate (%)",
      fill = "Agreement %"
    ) +
    theme_minimal()
}
```

```{r field-comparison2, eval=TRUE}


if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  
  # Prepare matched subsets, keep datasetName
  tb1 <- table1_matched %>%
    filter(match_status == "matched") %>%
    select(
      datasetName,        # keep for stratification
      match_id,
      any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
    ) %>%
    arrange(match_id)
  
  tb2 <- table2 %>%
    filter(id_CC %in% tb1$match_id) %>%
    arrange(id_CC)
  
  # Ensure same order
  tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]
  
  # Columns to compare
  comparison_cols <- intersect(names(tb1), names(tb2))
  comparison_cols <- setdiff(comparison_cols, c("match_id", "datasetName"))
  
  # Logical matrix of equalities for each field
  equal_df <- purrr::map_dfc(
    comparison_cols,
    ~ tb1[[.x]] == tb2[[.x]]
  )
  names(equal_df) <- comparison_cols
  
  # Long format: one row per (datasetName, field, row-pair)
  agreement_summary <- bind_cols(
    datasetName = tb1$datasetName,
    equal_df
  ) %>%
    pivot_longer(
      cols = -datasetName,
      names_to = "field",
      values_to = "equal"
    ) %>%
    group_by(datasetName, field) %>%
    summarise(
      agreements     = sum(equal, na.rm = TRUE),
      total          = sum(!is.na(equal)),
      agreement_rate = round(100 * agreements / total, 1),
      .groups        = "drop"
    ) %>%
    arrange(datasetName, desc(agreement_rate))
  
  kable(agreement_summary, caption = "Field Agreement Rates for Matched Rows, by datasetName")
  
} else {
  cat("No matched rows to compare.\n")
}
```





```{r agreement-plot2, fig.cap="Field Agreement Rates by datasetName"}

if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  agreement_summary %>%
    mutate(field = forcats::fct_reorder(field, agreement_rate)) %>%
    ggplot(aes(x = field, y = agreement_rate, fill = agreement_rate)) +
    geom_col() +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    coord_flip() +
    scale_fill_gradient(low = "orange", high = "darkgreen") +
    facet_wrap(~ datasetName) +
    labs(
      title = "Field Agreement Rates by datasetName",
      subtitle = "Comparison between matched table1 and table2 rows",
      x = "Field",
      y = "Agreement Rate (%)",
      fill = "Agreement %"
    ) +
    theme_minimal()
}

```

# Export Results

## Matched Records

```{r export-matched}
table1_export <- table1_matched %>%
  left_join(
    table1_raw %>% select(id_CC, starts_with(".")),
    by = "id_CC"
  ) %>%
  # select(
  #   id_CC_original = id_CC,
  #   match_id,
  #   match_status,
  #   match_quality,
  #   n_keys_used,
  #   n_core_keys_used,
  #   n_candidates_final,
  #   match_keys_used,
  #   datasetName,
  #   canonicalName,
  #   year,
  #   eventDate,
  #   locality,
  #   decimalLatitude,
  #   decimalLongitude,
  #   bibliographicCitation,
  #   catalogNumber,
  #   occurrenceID,
  #   institutionCode,
  #   Duplicate,
  #   starts_with(".")
  # ) %>%
  arrange(match_status, match_quality, datasetName, id_CC)

write.csv(
  table1_export,
  "./outputs/Table1_matched_strict_Argentina.csv",
  fileEncoding = "UTF-8",
  row.names = FALSE
)

cat("Exported", nrow(table1_export), "rows to Table1_matched_strict.csv\n")
```

## Good-to-bad  match ratio  

```{r}
write.csv(
    good.and.bad.matches,
    "./outputs/Table1_good-bad_matches_Argentina.csv",
    fileEncoding = "UTF-8", row.names = FALSE, sep=","
         )
```

## Unmatched Records for Manual Review

```{r export-unmatched, eval=FALSE}
if (exists("unmatched_review") && nrow(unmatched_review) > 0) {
  write.csv(
    unmatched_review,
    "./outputs/Table1_unmatched_review.csv",
    fileEncoding = "UTF-8",
    row.names = FALSE
  )
  
  cat("Exported", nrow(unmatched_review), "unmatched rows for review\n")
}
```

## Ambiguous Matches for Investigation

```{r export-ambiguous, eval=FALSE}
ambiguous_export <- table1_matched %>%
  filter(match_status == "ambiguous") %>%
  select(
    id_CC,
    n_candidates_final,
    n_keys_used,
    match_keys_used,
    datasetName,
    canonicalName,
    year,
    locality,
    decimalLatitude,
    decimalLongitude,
    bibliographicCitation,
    catalogNumber,
    occurrenceID,
    institutionCode
  ) %>%
  arrange(desc(n_candidates_final), datasetName)

if (nrow(ambiguous_export) > 0) {
  write.csv(
    ambiguous_export,
    "./outputs/Table1_ambiguous_matches.csv",
    fileEncoding = "UTF-8",
    row.names = FALSE
  )
  
  cat("Exported", nrow(ambiguous_export), "ambiguous matches for investigation\n")
}
```

# Summary Statistics Table

```{r summary-table}
summary_table <- data.frame(
  Metric = c(
    "Total table1 rows",
    "Successfully matched",
    "Ambiguous (multiple candidates)",
    "No match found",
    "No keys available",
    "Match rate (%)",
    "One-to-one constraint satisfied"
  ),
  Value = c(
    nrow(table1_matched),
    sum(table1_matched$match_status == "matched", na.rm = TRUE),
    sum(table1_matched$match_status == "ambiguous", na.rm = TRUE),
    sum(table1_matched$match_status == "no_match", na.rm = TRUE),
    sum(table1_matched$match_status == "no_keys_available", na.rm = TRUE),
    round(100 * sum(table1_matched$match_status == "matched", na.rm = TRUE) / nrow(table1_matched), 1),
    ifelse(nrow(duplicate_matches) == 0, "Yes ", "No ")
  )
)

kable(summary_table, caption = "Final Summary Statistics")
```



```{r}
# 
# table1_export[,c("id_CC_original", "match_id", "datasetName", "canonicalName", "year", 
# "locality", "decimalLatitude", "decimalLongitude", "bibliographicCitation", 
# "occurrenceID", "catalogNumber", "institutionCode")] %>% 

```



# Session Information

```{r session-info}
sessionInfo()
```

# Conclusion

This strictly conservative matching strategy ensures maximum reliability by:

1. **No tracking of matched table2 rows**: Each table1 row is evaluated independently against the entire table2
2. **Incremental key addition**: Keys are added one-by-one until uniqueness is achieved
3. **Explicit ambiguity detection**: Rows with multiple candidates are flagged as ambiguous
4. **Complete transparency**: All key combinations used are recorded for validation

**Final Results:**

- **Total processed:** `r nrow(table1_matched)` rows
- **Successfully matched:** `r sum(table1_matched$match_status == "matched", na.rm = TRUE)` rows
- **Match rate:** `r round(100 * sum(table1_matched$match_status == "matched", na.rm = TRUE) / nrow(table1_matched), 1)`%
- **One-to-one verified:** `r ifelse(nrow(duplicate_matches) == 0, " Yes", " No")`

All matched records represent unambiguous, unique links between table1 and table2.
