---
title: "One-to-One Record Matching Without Shared Keys: Integrating Manual Review into Updated Database Version"
author: "Biodiversity Data Integration"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
geometry: margin=2cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

# Introduction  

Durante el proceso de consolidación y depuración de datos, numerosas ocurrencias duplicadas (registros provenientes de una misma fuente e incorporados en distintos conjuntos de datos) no fueron detectadas por el algoritmo para de-duplicar datos, y debieron identificarse manualmente mediante planillas y QGIS. Asimismo, se realizaron mejoras puntuales en muchos registros, corrigiendo errores, completando información faltante, precisando coordenadas, añadiendo referencias asociadas, entre otros. Todas estas modificaciones se documentaron en la planilla `AtlasIctio_COMMENTS_2025_12.csv`.  

En el caso de los duplicados, la columna `Duplicate` (no confundir con `duplicated`) registra dichas evaluaciones: los valores 1 y 0 indican ocurrencias a retener o excluir, respectivamente, mientras que las celdas vacías representan registros únicos. En la columna `Comment` se incluyeron comentarios diversos  la mayoría de los cuales fueron pasados a otras columnas con el prefijo "." (por ejemplo, `.decimalLatitude`), donde un valor distinto de vacío indica el valor corregido del campo original correspondiente. `Comment` también contiene la palabra clave "remover" que indica  que hay que remover la fila. 

No obstante, el trabajo realizado sobre distintas versiones de la base consolidada provocó la pérdida del vínculo directo entre la tabla principal depurada y la tabla con anotaciones. Por ello, fue necesario restablecerlo mediante un protocolo de comparación exhaustivo entre tablas, implementando herramientas diagnósticas y verificaciones manuales en los casos dudosos.  

Este documento describe el método empleado para reestablecer el vínculo uno-a-uno entre ambas tablas, asegurando correspondencias precisas y completamente trazables.  

**Principios clave del flujo de trabajo**:  

  - Cada fila de `table1` debe coincidir de manera exacta con una única fila de `table2.`  
  - Las comparaciones se realizan exhaustivamente contra todas las filas de la otra tabla.  
  - Las coincidencias se establecen mediante combinaciones incrementales de campos clave previamente armonizados.  
  - El orden estratégico de estos campos favorece coincidencias correctas, conservadoras y no ambiguas.   

# Load Required Libraries

```{r libraries}

setwd("C:/Users/crist/Documents/GitHub/Atlas_Ictiogeografico/table_joining")

rm(list=ls())
  
library(dplyr)
library(stringr)
library(ggplot2)
library(knitr)
library(tidyr)
```


# Data importation  

```{r data-entry}
# Load data to start from here 

X.filtered.dedup <- readRDS("./input/dataset_filtered_deduplicated_before_manual_corrections_2026-05-20.rds")

```


Replace current datasetName by the old datasetName that was stored in datasetName_tmp.

```{r}
X.filtered.dedup <- X.filtered.dedup %>% mutate(datasetName = datasetName_tmp)
```


```{r}
CM <- read.csv("./input/AtlasIctio_COMMENTS_2025_12.csv", encoding = "UTF-8", sep = ",", colClasses = "character", na.strings = c("", "NA"))

```


<!-- En esta tabla se excluyeron algunos bloques de datos desde la planilla: -->
<!--   - (DatasetName == GBIF) & (institutionCode == iNaturalist). Ahora los datos de iNat tienen precedencia, y además se excluyenos los datos de iNat vía GBIF. -->
<!--   - (datasetName == "gbif") & (str_sub(occurrenceID, 1L, 3L) == "mma"). Usamos la versión mejorada directamente del MMA, no la versión  publicada en GBIF. -->
<!--   - (datasetName == "subpesca-1") una lista de papers indexados en otras fuentes fue excluido. Ver planilla en la respectiva carpeta,  y código de exclusión arriba. -->


First, fix truncated column names such that they match those of the database: 

```{r}
# Restore full column names from Maryse's table

full_names  <- names(X.filtered.dedup)
short_names <- names(CM)

new_names <- sapply(short_names, function(s) {
  # try exact prefix match: full name starts with the short name
  cand <- full_names[substr(full_names, 1, nchar(s)) == s]
  if (length(cand) == 1) {
    cand          # unique match → use full name
  } else {
    s             # no unique match → keep original
  }
}, USE.NAMES = FALSE)
# Check
# data.frame(old_name = colnames(CM),
#            full_name = new_names)
names(CM) <- new_names

CM <- CM %>%
  rename(
    datasetName     = datasetNam,
    institutionID   = institut_1,
    institutionCode = institutio,
    occurrenceID    = occurrence,
    occurrenceStatus = occurren_1,
    occurrenceRemarks = occurren_2)

# sort(colnames(CM))
# sort(colnames(X))
# CM %>% select(occurren_1, occurren_2) %>% distinct()
# 
# X %>% select(occurrenceRemarks, occurrenceStatus) %>% distinct()
rm("full_names", "new_names", "short_names")

```




Merge (coalesce) two columns that were separated only due to a typo.  
```{r}
# Restore info from bibiographicCitation info

CM <- CM %>%
  mutate(
    across(c(bibliographicCitation, bibiograph),
           ~ str_trim(.) %>% na_if("") %>% na_if("NA")),
    bibliographicCitation = coalesce(bibliographicCitation, bibiograph)
  ) %>%  select(-bibiograph)   

```


# Data Preparation  

## Filter Source Data  

Select columns from the comments table (CM) that can be used to match rows against the master databese, and all rows that were annotated requiring an action:

```{r prepare-table1}


# Table1: rows with changes or exclusion requests
# Duplicate: "0" = exclude, "1" = keep, NA = no action needed
# Columns with "." prefix contain proposed changes
table1_raw <- CM %>% 
  filter(
    Duplicate == "0" | 
    !is.na(Comment) | 
    if_any(starts_with("."), ~ !is.na(.x))
  ) %>%
  select(
    id_CC,
    Duplicate,
    decimalLatitude,
    decimalLongitude,
    canonicalName, 
    year, 
    eventDate, 
    locality, 
    samplingProtocol,
    individualCount,
    catalogNumber, 
    datasetName,
    bibliographicCitation,
    institutionCode,
    occurrenceID, 
    Comment,
    starts_with(".")
  )

cat("Table1 rows requiring action:", nrow(table1_raw), "\n")
```

Note: There were several datasets that are not carried forward because no changes needed to be made. These were:  

```{r}
setdiff( sort(unique(CM$datasetName)), sort(unique(table1_raw$datasetName)))

```


```{r prepare-table2}
# Table2: consolidated master table after de-duplication
table2_raw <- X.filtered.dedup %>%
  select(any_of(c("id_CC", "occurrenceID",colnames(table1_raw))))

cat("Table2 total rows:", nrow(table2_raw), "\n")

# Verify id_CC is unique
if (any(duplicated(table2_raw$id_CC))) {
  stop("ERROR: id_CC in table2 is not unique!")
}
```

## Standardization Function

Reduce meaningless variation in key fields:

```{r standardization-function}
standardize_for_matching <- function(df) {
  df %>%
    mutate(
      # Coordinates: ensure numeric (no rounding to preserve precision)
      decimalLatitude  = as.numeric(decimalLatitude),
      decimalLongitude = as.numeric(decimalLongitude),
      
      # Event date: extract YYYY-MM-DD format
      eventDate = if_else(
        is.na(eventDate),
        NA_character_,
        str_sub(as.character(eventDate), 1L, 10L)
      ),
      
      # Year: from year column or extract from eventDate
      year = {
        y1 <- suppressWarnings(as.integer(year))
        y2 <- suppressWarnings(as.integer(str_extract(eventDate, "\\d{4}")))
        coalesce(y1, y2)
      },
      
      # Bibliographic citation: first 20 chars (avoid minor variations)
      bibliographicCitation = str_sub(
        str_trim(as.character(bibliographicCitation)),
        1, 20
      ),
      
      # Text fields: trim, lowercase, standardize NAs
      across(
        c(canonicalName, locality, datasetName, catalogNumber,
          occurrenceID, institutionCode),
        ~ {
          x <- str_to_lower(str_trim(as.character(.x)))
          x[x %in% c("", "na", "n a", "n/a")] <- NA_character_
          x
        }
      ),
      
      # Dataset name: standardize known variations
      datasetName = case_when(
        datasetName %in% c("subpesca-1", "subpesca-2",
                           "fish database correa-boisjoly",
                           "fishgis_metadatabase_ccorrea") ~ datasetName,
        TRUE ~ str_sub(datasetName, 1L, 4L)
      )
    )
}
```

Create a lookup table for occurrenceID from the master table for later use
```{r}
occurrenceID_lookup <- table2_raw %>%
  select(id_CC, occurrenceID_master = occurrenceID)
```


## Apply Standardization

```{r apply-standardization}
table1 <- standardize_for_matching(table1_raw)
table2 <- standardize_for_matching(table2_raw)

cat("Standardization complete.\n")
cat("Table1:", nrow(table1), "rows\n")
cat("Table2:", nrow(table2), "rows\n")
```

# Matching Strategy

## Define Matching Keys

Keys are ordered by reliability:

```{r define-keys}
# Core keys: most reliable identifiers
core_keys <- c(
  "datasetName",
  "canonicalName",
  "decimalLatitude",
  "decimalLongitude",
  "bibliographicCitation")

# Extra keys: used to resolve ambiguities
extra_keys <- c(
  "catalogNumber",
  "year",
  "locality",
  "samplingProtocol",
  "individualCount",
  "eventDate",
  "institutionCode",
  "occurrenceID")

all_keys <- c(core_keys, extra_keys)

cat("Matching will use up to", length(all_keys), "keys in hierarchical order\n")
cat("Core keys:", length(core_keys), "\n")
cat("Extra keys:", length(extra_keys), "\n")
```

## Strictly Conservative Matching Function

This function adds keys incrementally until a unique match is found. If no unique match is found, no assignment is  done, and the match_id field is left blank:

```{r matching-function}
match_tables_strict <- function(table1, table2, core_keys, extra_keys) {
  
  # Validate inputs
  if (!"id_CC" %in% names(table2)) {
    stop("table2 must contain unique identifier 'id_CC'")
  }
  
  if (any(duplicated(table2$id_CC))) {
    stop("id_CC in table2 must be unique")
  }
  
  # Initialize result
  result <- table1 %>%
    mutate(
      row_id = row_number(),
      match_id = NA_integer_,
      match_keys_used = NA_character_,
      n_keys_used = NA_integer_,
      n_candidates_final = NA_integer_,
      match_status = NA_character_
    )
  
  all_keys <- c(core_keys, extra_keys)
  n_core <- length(core_keys)
  
  # Process each row of table1
  for (i in seq_len(nrow(table1))) {
    
    row_i <- table1[i, , drop = FALSE]
    
    # Start with ALL rows from table2 (no tracking of matched rows)
    candidates <- table2
    
    keys_used <- character(0)
    keys_available <- character(0)
    
    # Identify which keys have non-NA values in this table1 row
    for (key in all_keys) {
      if (key %in% names(row_i) && !is.na(row_i[[key]])) {
        keys_available <- c(keys_available, key)
      }
    }
    
    # Incrementally add keys until unique match or keys exhausted
    for (key in keys_available) {
      
      val <- row_i[[key]]
      
      # Filter candidates: exact match on this key, non-NA in table2
      candidates <- candidates %>%
        filter(!is.na(.data[[key]]), .data[[key]] == val)
      
      keys_used <- c(keys_used, key)
      
      # Stop if no candidates remain
      if (nrow(candidates) == 0) {
        result$match_status[i] <- "no_match"
        result$n_candidates_final[i] <- 0
        break
      }
      
      # Stop if unique match achieved
      if (nrow(candidates) == 1) {
        result$match_id[i] <- candidates$id_CC[1]
        result$match_keys_used[i] <- paste(keys_used, collapse = ";")
        result$n_keys_used[i] <- length(keys_used)
        result$n_candidates_final[i] <- 1
        result$match_status[i] <- "matched"
        break
      }
    }
    
    # If we exhausted all keys without achieving uniqueness
    if (is.na(result$match_status[i]) && nrow(candidates) > 1) {
      result$match_status[i] <- "ambiguous"
      result$n_candidates_final[i] <- nrow(candidates)
      result$match_keys_used[i] <- paste(keys_used, collapse = ";")
      result$n_keys_used[i] <- length(keys_used)
    }
    
    # If no keys were available for this row
    if (length(keys_available) == 0) {
      result$match_status[i] <- "no_keys_available"
      result$n_candidates_final[i] <- nrow(table2)
    }
  }
  
  # Classify match quality based on which keys were used
  result <- result %>%
    mutate(
      n_core_keys_used = if_else(
        !is.na(match_keys_used),
        sapply(strsplit(match_keys_used, ";"), function(keys) {
          sum(keys %in% core_keys)
        }),
        NA_integer_
      ),
      match_quality = case_when(
        match_status != "matched" ~ match_status,
        n_core_keys_used == n_core ~ "all_core_firlds",
        n_core_keys_used >= 4 ~ "4plus_core",
        n_core_keys_used == 3 ~ "3_core",
        n_core_keys_used == 2 ~ "2_core",
        n_core_keys_used <= 1 ~ "1_core",
        TRUE ~ "matched"
      )
    )
  
  result %>% select(-row_id)
}
```

## Execute Matching

```{r execute-matching}
cat("Starting strictly conservative matching...\n")
cat("Each table1 row will be matched independently against ALL table2 rows.\n\n")

table1_matched <- match_tables_strict(
  table1 = table1,
  table2 = table2,
  core_keys = core_keys,
  extra_keys = extra_keys
)

cat("Matching complete.\n")
```

# Results and Diagnostics

## Overall Matching Success

```{r matching-summary}
summary_stats <- table1_matched %>%
  summarise(
    total_rows = n(),
    matched = sum(match_status == "matched", na.rm = TRUE),
    ambiguous = sum(match_status == "ambiguous", na.rm = TRUE),
    no_match = sum(match_status == "no_match", na.rm = TRUE),
    no_keys = sum(match_status == "no_keys_available", na.rm = TRUE),
    match_rate = round(100 * matched / total_rows, 1)
  )

kable(summary_stats, caption = "Overall Matching Results")
```

## Match Status Distribution

```{r match-status}
status_summary <- table1_matched %>%
  count(match_status) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1)
  ) %>%
  arrange(desc(n))

kable(status_summary, caption = "Match Status Breakdown")
```



## Match Quality for Successfully Matched Rows

```{r match-quality}
quality_summary <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_quality) %>%
  mutate(
    percentage = round(100 * n / sum(n), 1)
  ) %>%
  arrange(desc(n))

kable(quality_summary, caption = "Match Quality Breakdown (Matched Rows Only)")
```

```{r quality-plot, fig.cap="Distribution of Match Quality"}
table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_quality) %>%
  mutate(match_quality = forcats::fct_reorder(match_quality, n)) %>%
  ggplot(aes(x = match_quality, y = n)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Match Quality Distribution",
    subtitle = "For successfully matched rows only",
    x = "Quality Category",
    y = "Number of Matches"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Keys Required for Matching

```{r keys-used}
keys_summary <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(n_keys_used, n_core_keys_used) %>%
  arrange(n_keys_used, desc(n_core_keys_used))

kable(keys_summary, caption = "Number of Keys Required to Achieve Uniqueness")
```


## Specific Key Combinations

Top 20 most common key combinations used:

```{r key-combinations}
key_combos <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_keys_used, sort = TRUE) %>%
  head(20)

kable(key_combos, caption = "Most Common Key Combinations")
```

## Uniqueness Validation

Critical check: verify each match_id appears only once:

```{r uniqueness-check}
duplicate_matches <- table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_id, name = "frequency") %>%
  filter(frequency > 1)

if (nrow(duplicate_matches) > 0) {
  cat(" ERROR:", nrow(duplicate_matches), "table2 rows matched multiple times!\n")
  kable(head(duplicate_matches, 20), caption = "Duplicate Matches (CRITICAL ERROR)")
} else {
  cat(" SUCCESS: All matches are unique (one-to-one)\n")
  cat("Each match_id appears exactly once.\n")
}
```

```{r uniqueness-plot, fig.cap="Match ID Frequency Distribution"}
table1_matched %>%
  filter(match_status == "matched") %>%
  count(match_id, name = "freq") %>%
  ggplot(aes(x = freq)) +
  geom_histogram(binwidth = 1, color = "black", fill = "steelblue") +
  geom_vline(xintercept = 1, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 1.5, y = Inf, label = "All should be at 1", 
           vjust = 2, color = "red", fontface = "bold") +
  labs(
    title = "Distribution of match_id Frequency",
    subtitle = "Perfect one-to-one matching requires all frequencies = 1",
    x = "Times each match_id appears",
    y = "Count"
  ) +
  theme_minimal()
```

## Ambiguous Matches Analysis

Rows that had multiple candidates even after using all available keys:

```{r ambiguous-analysis}
if (sum(table1_matched$match_status == "ambiguous", na.rm = TRUE) > 0) {
  
  ambiguous_summary <- table1_matched %>%
    filter(match_status == "ambiguous") %>%
    group_by(n_candidates_final) %>%
    summarise(
      count = n(),
      avg_keys_used = round(mean(n_keys_used, na.rm = TRUE), 1)
    ) %>%
    arrange(desc(count))
  
  kable(ambiguous_summary, caption = "Ambiguous Matches: Number of Final Candidates")

# Show examples of ambiguous matches
  ambiguous_examples <- table1_matched %>%
    filter(match_status == "ambiguous") %>%
    select(
      id_CC, occurrenceID, datasetName, canonicalName, year, 
      n_candidates_final, n_keys_used, match_keys_used
    ) %>%
    head(10)
  
  kable(ambiguous_examples, caption = "Examples of Ambiguous Matches")
  
} else {
  cat("No ambiguous matches found.\n")
}
```

## No Match Analysis

Rows where no table2 row matched the criteria:

```{r no-match-analysis}
if (sum(table1_matched$match_status == "no_match", na.rm = TRUE) > 0) {
  
  no_match_summary <- table1_matched %>%
    filter(match_status == "no_match") %>%
    count(datasetName) %>%
    arrange(desc(n))
  
  kable(no_match_summary, caption = "No Match Cases by Dataset")
  
} else {
  cat("All rows either matched or were ambiguous.\n")
}
```

## Unmatched Rows Requiring Review

Exclude expected unmatched categories:


```{r unmatched-review}
unmatched_review <- table1_matched %>%
  filter(
     match_status != "matched",
        ) %>%
  arrange(match_status, datasetName, occurrenceID, canonicalName, year) 

cat("Unmatched rows requiring manual review:", nrow(unmatched_review), "\n\n")

unmatched_by_status <- unmatched_review %>%
  count(match_status, datasetName) %>%
  arrange(match_status, desc(n))

kable(unmatched_by_status, caption = "Unmatched Rows by Status and Dataset")
```

## Matched, rowwise field-by-Field Comparison  

For every matched row, comare pairs of corresponding fields, and anayze the frequencies of matching and miss-matching filds, and the log ratio of matches to miss-matches.     

```{r rowwise-field-comparison, eval=TRUE} 

if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {

# matched subsets
tb1 <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    match_id, occurrenceID,
    any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
  ) %>%
  arrange(match_id)

tb2 <- table2 %>%
  filter(id_CC %in% tb1$match_id) %>%
  arrange(id_CC)

tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]

comparison_cols <- setdiff(intersect(names(tb1), names(tb2)), "match_id")

diff.strict <- tb1[, comparison_cols] == tb2[, comparison_cols]

tmp <- table1_matched %>%
  filter(match_status == "matched") %>%
  left_join(occurrenceID_lookup, by = c("match_id" = "id_CC")) %>%
  select(
    id_CC_old = id_CC,
    match_id,
    occurrenceID_master,
    match_keys_used,
    dataset = datasetName
  ) %>%
  arrange(match_id)

# logical match matrix -> data frame
diff.strict <- as.data.frame(diff.strict)

# add metadata for each matched row
diff.strict <- bind_cols(
  tmp[!is.na(tmp$match_id), ],  # keep only matched rows in tmp
  diff.strict
)

# compute number of matching fields row-wise
diff.strict <- diff.strict %>%
  mutate(fieldgood = rowSums(across(!c(id_CC_old, match_id, occurrenceID_master,
                                       match_keys_used, dataset)), na.rm = TRUE))

# plot
ggplot(diff.strict, aes(x = dataset, y = fieldgood)) +
  geom_jitter(aes(color = fieldgood),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
                     limits= c(1, length(comparison_cols)),
                     expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Row-wise Agreement: Number of Fields Matched",
    x = NULL,
    y = "Count of matching fields"
  ) +
  theme_minimal()

} else {
  cat("No matched rows to compare.\n")
}

```

```{r rowwise-field-comparison2, eval=TRUE} 


 if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {

# matched subsets
tb1 <- table1_matched %>%
  filter(match_status == "matched") %>%
  select(
    match_id, occurrenceID,
    any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
  ) %>%
  arrange(match_id)

tb2 <- table2 %>%
  filter(id_CC %in% tb1$match_id) %>%
  arrange(id_CC)

tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]

comparison_cols <- setdiff(intersect(names(tb1), names(tb2)), "match_id")

# COUNT DIFFERENCES THIS TIME
diff.strict2 <- tb1[, comparison_cols] != tb2[, comparison_cols]

tmp <- table1_matched %>%
  filter(match_status == "matched") %>%
  left_join(occurrenceID_lookup, by = c("match_id" = "id_CC")) %>%
  select(
    id_CC_old = id_CC,
    match_id,
    occurrenceID_master,
    match_keys_used,
    dataset = datasetName
  ) %>%
  arrange(match_id)

# logical match matrix -> data frame
diff.strict2 <- as.data.frame(diff.strict2)

# add metadata for each matched row
diff.strict2 <- bind_cols(
  tmp[!is.na(tmp$match_id), ],  # keep only matched rows in tmp
  diff.strict2
)

# compute number of matching fields row-wise
diff.strict.falses <- diff.strict2 %>%
  mutate(fieldbad = rowSums(across(!c(id_CC_old, match_id,occurrenceID_master, match_keys_used, dataset)), na.rm = TRUE))

# plot
ggplot(diff.strict.falses, aes(x = dataset, y = fieldbad)) +
  geom_jitter(aes(color = fieldbad),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
                     limits= c(1, length(comparison_cols)),
                     expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  labs(
    title = "Row-wise disagreement: Number of unmatching fields",
    x = NULL,
    y = "Count of non-matching fields"
  ) +
  theme_minimal()

} else {
  cat("No matched rows to compare.\n")
}



```


```{r}

good.and.bad.matches <-  diff.strict %>% full_join(diff.strict.falses, by = "id_CC_old", keep = T) %>% 
  select(dataset.x, id_CC_old.x, 
         occurrenceID_master.x,
         match_id.x, 
         fieldgood, fieldbad) %>% 
  mutate(logodd.ratio = log10((fieldgood + 1) / (fieldbad + 1)), 
         precent.good = 100 * (fieldgood /length(comparison_cols)),
         percent.bad = 100 *  (fieldbad /length(comparison_cols))) 

good.and.bad.matches %>% filter(fieldbad >= 3) %>% 
  kable(caption = "Matched rows with 3 or more missmatches.")

```

```{r}
good.and.bad.matches %>% 
ggplot(aes(x = dataset.x, y = logodd.ratio)) +
  geom_jitter(aes(color = logodd.ratio),
              height = 0.25, width = 0.25, alpha = 0.5) +
  coord_flip() +
  # scale_y_continuous(breaks = seq(1, length(comparison_cols)), 
  #                    limits= c(1, length(comparison_cols)),
  #                    expand = FALSE) +
  scale_color_viridis_c(option = "plasma", direction = -1) +
  geom_hline(yintercept = 0, lty=2, col=2)+
  labs(
    title = "Row-wise Agreement: Log-Ratio of matched-to-unmatched fields",
    x = NULL,
    y = "Log of the odds"
  ) +
  theme_minimal()


```



## Matched, overall field-by-Field Comparison  

For successfully matched rows, compare original values:

```{r overall-field-comparison, eval=TRUE}
 if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  
  # Prepare matched subsets
  tb1 <- table1_matched %>%
    filter(match_status == "matched") %>%
    select(
      match_id,
      any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
    ) %>%
    arrange(match_id)
  
  tb2 <- table2 %>%
    filter(id_CC %in% tb1$match_id) %>%
    arrange(id_CC)
  
  # Ensure same order
  tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]
  
  # Compare field by field
  comparison_cols <- intersect(names(tb1), names(tb2))
  comparison_cols <- setdiff(comparison_cols, "match_id")
  
  agreement <- sapply(comparison_cols, function(col) {
    sum(tb1[[col]] == tb2[[col]], na.rm = TRUE)
  })
  
  total <- nrow(tb1)
  
  agreement_summary <- data.frame(
    field = comparison_cols,
    agreements = agreement,
    total = total,
    agreement_rate = round(100 * agreement / total, 1)
  ) %>%
    arrange(desc(agreement_rate))
  
  kable(agreement_summary, caption = "Field Agreement Rates for Matched Rows")
  
} else {
  cat("No matched rows to compare.\n")
}
```



```{r agreement-plot, eval=F, fig.cap="Field Agreement Rates"}
if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  agreement_summary %>%
    mutate(field = forcats::fct_reorder(field, agreement_rate)) %>%
    ggplot(aes(x = field, y = agreement_rate, fill = agreement_rate)) +
    geom_col() +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    coord_flip() +
    scale_fill_gradient(low = "orange", high = "darkgreen") +
    labs(
      title = "Field Agreement Rates",
      subtitle = "Comparison between matched table1 and table2 rows",
      x = "Field",
      y = "Agreement Rate (%)",
      fill = "Agreement %"
    ) +
    theme_minimal()
}
```

```{r field-comparison2, eval=TRUE}


if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  
  # Prepare matched subsets, keep datasetName
  tb1 <- table1_matched %>%
    filter(match_status == "matched") %>%
    select(
      datasetName,        # keep for stratification
      match_id,
      any_of(setdiff(colnames(table2), c("id_CC", starts_with("."))))
    ) %>%
    arrange(match_id)
  
  tb2 <- table2 %>%
    filter(id_CC %in% tb1$match_id) %>%
    arrange(id_CC)
  
  # Ensure same order
  tb2 <- tb2[match(tb1$match_id, tb2$id_CC), ]
  
  # Columns to compare
  comparison_cols <- intersect(names(tb1), names(tb2))
  comparison_cols <- setdiff(comparison_cols, c("match_id", "datasetName"))
  
  # Logical matrix of equalities for each field
  equal_df <- purrr::map_dfc(
    comparison_cols,
    ~ tb1[[.x]] == tb2[[.x]]
  )
  names(equal_df) <- comparison_cols
  
  # Long format: one row per (datasetName, field, row-pair)
  agreement_summary <- bind_cols(
    datasetName = tb1$datasetName,
    equal_df
  ) %>%
    pivot_longer(
      cols = -datasetName,
      names_to = "field",
      values_to = "equal"
    ) %>%
    group_by(datasetName, field) %>%
    summarise(
      agreements     = sum(equal, na.rm = TRUE),
      total          = sum(!is.na(equal)),
      agreement_rate = round(100 * agreements / total, 1),
      .groups        = "drop"
    ) %>%
    arrange(datasetName, desc(agreement_rate))
  
  kable(agreement_summary, caption = "Field Agreement Rates for Matched Rows, by datasetName")
  
} else {
  cat("No matched rows to compare.\n")
}
```





```{r agreement-plot2, fig.cap="Field Agreement Rates by datasetName"}

if (sum(table1_matched$match_status == "matched", na.rm = TRUE) > 0) {
  agreement_summary %>%
    mutate(field = forcats::fct_reorder(field, agreement_rate)) %>%
    ggplot(aes(x = field, y = agreement_rate, fill = agreement_rate)) +
    geom_col() +
    geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
    coord_flip() +
    scale_fill_gradient(low = "orange", high = "darkgreen") +
    facet_wrap(~ datasetName) +
    labs(
      title = "Field Agreement Rates by datasetName",
      subtitle = "Comparison between matched table1 and table2 rows",
      x = "Field",
      y = "Agreement Rate (%)",
      fill = "Agreement %"
    ) +
    theme_minimal()
}

```

# Export Results

## Matched Records

Import preious version that was manually checked, just to help flag problemetic cases. 
```{r}
legacy_manual_check <- read.csv("Cambios_a_implementar_2025-12-15_TMP.csv") %>% 
  select(id_CC_original, correcting_match) %>% 
  mutate(id_CC_original = as.character(id_CC_original),  
         correcting_match_old = correcting_match, 
         .keep = "unused") %>% 
  mutate(correcting_match = as.character(NA)) %>% relocate(correcting_match)

```

Prepare diagnostic infornation to add it to the overall table for revision
```{r}
good.to.bad <- good.and.bad.matches %>% rename(id_CC_original = id_CC_old.x ) %>% 
  select(id_CC_original, 
         fieldgood, fieldbad, 
         logodd.ratio,
         precent.good,
         percent.bad) 
```


```{r finalResultsTable}
table1_export <- table1_matched %>% 
 mutate(id_CC_original = id_CC, .keep = "unused") %>% 
 left_join(legacy_manual_check, by = "id_CC_original") %>% 
 left_join(occurrenceID_lookup, by = c("match_id" = "id_CC")) %>%
  mutate(occ_match = occurrenceID == str_to_lower(str_trim(as.character(occurrenceID_master)))) %>% 
 left_join(good.to.bad, by = "id_CC_original") %>%
  select(
    id_CC_original,
    match_id,
    correcting_match,
    correcting_match_old,
    occurrenceID_master,
    occurrenceID,
    occ_match,
    match_status,
    match_quality,
    fieldgood, 
    fieldbad, 
    logodd.ratio,
    precent.good,
    percent.bad,
    n_keys_used,
    n_core_keys_used,
    n_candidates_final,
    match_keys_used,
    datasetName,
    canonicalName,
    year,
    eventDate,
    locality,
    decimalLatitude,
    decimalLongitude,
    bibliographicCitation,
    catalogNumber,
    institutionCode,
    Duplicate,
    Comment,
    starts_with(".")
  ) %>%
  arrange(as.numeric(id_CC_original)) #%>% 
  #arrange(match_status, match_quality, datasetName, canonicalName, year)


```


Export table to hard drive for manual revision
```{r export-matched}
write.csv( table1_export,
  "./outputs/Table1_matched_strict_2026-05-26.csv",
  fileEncoding = "UTF-8", row.names = FALSE )

# But see Utility function at the end to import previos annotations, if available. 

cat("Exported", nrow(table1_export), "rows to Table1_matched_strict.csv\n")
```



# Summary Statistics Table

```{r summary-table}
summary_table <- data.frame(
  Metric = c(
    "Total table1 rows",
    "Successfully matched",
    "Ambiguous (multiple candidates)",
    "No match found",
    "No keys available",
    "Match rate (%)",
    "One-to-one constraint satisfied"
  ),
  Value = c(
    nrow(table1_matched),
    sum(table1_matched$match_status == "matched", na.rm = TRUE),
    sum(table1_matched$match_status == "ambiguous", na.rm = TRUE),
    sum(table1_matched$match_status == "no_match", na.rm = TRUE),
    sum(table1_matched$match_status == "no_keys_available", na.rm = TRUE),
    round(100 * sum(table1_matched$match_status == "matched", na.rm = TRUE) / nrow(table1_matched), 1),
    ifelse(nrow(duplicate_matches) == 0, "Yes ", "No ")
  )
)

kable(summary_table, caption = "Final Summary Statistics")
```


# Final remarks    

- **Total processed:** `r nrow(table1_matched)` rows  
- **Successfully matched:** `r sum(table1_matched$match_status == "matched", na.rm = TRUE)` rows  
- **Match rate:** `r round(100 * sum(table1_matched$match_status == "matched", na.rm = TRUE) / nrow(table1_matched), 1)`%  
- **One-to-one verified:** `r ifelse(nrow(duplicate_matches) == 0, " Yes", " No")`  

This strictly conservative matching strategy ensures maximum reliability by:  

1. **No tracking of matched table2 rows**: Each table1 row is evaluated independently against the entire table2  
2. **Incremental key addition**: Keys are added one-by-one until uniqueness is achieved  
3. **Explicit ambiguity detection**: Rows with multiple candidates are flagged as ambiguous  
4. **Complete transparency**: All key combinations used are recorded for validation  


## Final Outputs

The automated record-matching workflow successfully identified most correct links between the master consolidated dataset and the AtlasIctio corrections table. Records flagged as unmatched, low-confidence, or ambiguous were manually checked to ensure traceability and accuracy.  

All verification results are now compiled in `Table1_matched_strict_reviewed_2026-05-26.csv`. This file establishes the mapping between:  

  - `dataset_before_Remocion_manual_2025-12-03.rds` (consolidated dataset prior to manual review)  

  - `AtlasIctio_COMMENTS_2025_12.csv` (corrected annotations table)  

The reviewed file includes the following fields required for updating identifiers:  

  - `id_CC_original`: ID of the consolidated dataset row before manual revision  

  - `match_id`: ID in the deduplicated dataset (`dataset_filtered_deduplicated_before_manual_corrections_2026-05-20.rds`) determined by the matching workflow  

  - `correcting_match_old`: manually validated match (integer if corrected match_id, or explanatory note if text) of a previous version conducted in december 2025. Just for the records and helping flag/resolve problematic records.   
  - `correcting_match`: field for manual validation
  

## Next Steps

- Integrate the reviewed match table into the main data processing workflow  

- Before joining to the main table, replace outdated match_id values where manual correction exists  


# Session Information

```{r session-info}

sessionInfo()

```


```{r, eval=T, include=T}
knitr::knit_exit()
```


# Utility function to compare assignments and import manually revised annotation to new matching tables. 


```{r, eval = F, include=FALSE}

# =====================================================================
# import_annotations() — Import Reviewed Annotations by Composite Key
# =====================================================================
#
# This function helps migrate manually reviewed annotations (e.g., 
# match corrections) from an older version of a dataset (`table_old`) 
# to a newly generated version (`table_new`).
#
# It matches rows based on a customizable set of key columns 
# (default: id_CC_original + match_id) and imports the annotation 
# column (default: "correcting_match") into the new table.
#
# The function also adds diagnostic columns to the output for easy 
# filtering and manual review. These indicate whether each row was 
# previously reviewed, whether the annotation was successfully 
# imported, and whether the original ID or match has changed.
#
# Output:
# - `updated_table`: Copy of the new table with annotations and diagnostics
# - `diagnostics`: Summary stats and rows needing review
#
# Ideal for use in iterative data matching, de-duplication, and 
# validation workflows.


# Load utility functiuon to compare two versions of matching tables, the older with comments in correcting_match. See help and fields' definitins in the source code. 

source("./annotated-new_output_comparisson/compare_and_import_annotations.R")

# Load the (older) annotated matched table

table1.annotated <- read.csv("./outputs/Table1_matched_strict_reviewed_2025-12-06.csv", 
               encoding = "UTF-8", sep = ",", colClasses = "character", 
               na.strings = c("", "NA"))


# Load the new version lacking manual annotations, or use the object generated in this script.   
# Run the comparisson function and import the correcting_match filed into the new table only if the compossite  matching key is the same  (combination of id_CC_original and match_id)

  result <- import_annotations(
    table_old = table1.annotated,
    table_new = table1_export,
    key_columns = c("id_CC_original", "match_id"),
    annotation_column = "correcting_match"
  )
  
# Access results 
  
# Main updated table
updated_table <- result$updated_table
  
# # Diagnostics (list)
  # result$diagnostics$summary # needing manual review just means have never been checked 


# Write the updated table including diagnostic columns on the right  
write.csv(updated_table, "./annotated-new_output_comparisson/Table1_matched_strict_2025-12-10_reviewing.csv", fileEncoding = "UTF-8", row.names = F)

  
```

# Utility function to simplify the table with annotations after manual review.

This is to strip the table with annotations and corrections from columns related to the row-matching process, diagnostic matching columns, etc. The idea is to get a clean  table with corrections to parse to the main pipeline. 

It runs in three passes: 
(1) update `occurrenceID_master` using manually corrected matches,
(2) pick the best available match ID, and 
(3) remove unlinked rows and tidy the column layout.


```{r}
CM_fixed <- read.csv("./outputs/Table1_matched_strict_2026-05-26_editedMaryse.csv", 
         stringsAsFactors = F, 
         fileEncoding = "UTF-8")


CM_fixed <- CM_fixed %>%
  mutate(
    id_CC_key_num = suppressWarnings(as.numeric(correcting_match)),
    id_CC_key = if_else(
      !is.na(id_CC_key_num) & id_CC_key_num == as.integer(id_CC_key_num),
      as.integer(id_CC_key_num),
      NA_integer_
    )
  ) %>%
  left_join(
    X.filtered.dedup %>%
      select(id_CC, occurrenceID_from_master = occurrenceID),
    by = c("id_CC_key" = "id_CC")
  ) %>%
  mutate(
    occurrenceID_master = if_else(
      !is.na(id_CC_key) & !is.na(occurrenceID_from_master),
      occurrenceID_from_master,
      occurrenceID_master
    )
  ) %>%
  select(-id_CC_key_num, -id_CC_key, -occurrenceID_from_master)


CM_fixed <- CM_fixed %>%
  mutate(
    correcting_match_num = suppressWarnings(as.numeric(correcting_match)),
    match_ID_num = suppressWarnings(as.numeric(match_id)),

    correcting_match_int = if_else(
      !is.na(correcting_match_num) &
        correcting_match_num == as.integer(correcting_match_num),
      as.integer(correcting_match_num),
      NA_integer_
    ),

    match_ID_int = if_else(
      !is.na(match_ID_num) &
        match_ID_num == as.integer(match_ID_num),
      as.integer(match_ID_num),
      NA_integer_
    ),

    new_match_id = coalesce(correcting_match_int, match_ID_int)
  ) %>%
  select(
    -correcting_match_num,
    -match_ID_num,
    -correcting_match_int,
    -match_ID_int
  )

# Remove rows flagged with correcting_match == "unlink", because these cases are no longer included in the master databese 

CM_fixed <- CM_fixed %>%
  filter(!str_detect(correcting_match, "^unlink")) %>% 
  select(-id_CC_original, -match_id, -correcting_match) %>% 
  rename(id_CC = new_match_id,
         datasetName_tmp = datasetName) %>% relocate(id_CC) 

# Create  a new column to indicate cases that need  to be removed from the master table

CM_fixed <- CM_fixed %>% mutate(Remove_from_master = 
                                  ifelse(
        str_detect(Comment, "^Remove"), 
        "remove",NA)) %>% 
       relocate(Remove_from_master, .after = Comment) 


```


```{r}
write.csv(CM_fixed, "./outputs/Cambios_a_implementar_2026-05-28.csv", fileEncoding = "UTF-8", row.names = F)


```


```{r, eval=T, include=T}
knitr::knit_exit()
```



